---
title: "California Housing Dataset Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We will be examining the well known **California Housing dataset**

The columns are:

- `longitude`
- `latitude`
- `housing_median_age`
- `total_rooms`
- `total_bedrooms`
- `population`
- `households`
- `median_income`
- `ocean_proximity` *(categorical string data)*
- `median_house_value`

## 1. Importing and Exploring the Data

```{r}
library(tidyverse)

url <- "https://ewan.my/datasets/housing.csv"
df <- read_csv(url)
```

---

Let's take a quick look at the first few rows and summary statistics of the dataset.

```{r}
head(df)

summary(df)

glimpse(df)
```

---

## 2. Data Cleaning and Processing

### 2.1 Checking for Missing Values

We'll check if there are any missing values in the dataset, especially in columns like `total_bedrooms`.

```{r}
missing_values <- colSums(is.na(df))
cat("Missing values in each column:\n")
print(missing_values)
```

### 2.2 Handling Missing Values

For columns with missing values (for example, `total_bedrooms`), you might decide to fill them with the median value, drop those rows, or apply another strategy.

```{r}
if (missing_values["total_bedrooms"] > 0) {
  median_bedrooms <- median(df$total_bedrooms, na.rm = TRUE)
  df$total_bedrooms <- ifelse(is.na(df$total_bedrooms), median_bedrooms, df$total_bedrooms)
  cat("Filled missing values in 'total_bedrooms' with median:", median_bedrooms, "\n")
}
```

### 2.3 Converting Categorical Data

Since `ocean_proximity` is a string, you might want to encode it before running numerical analyses or machine learning algorithms. One common method is one-hot encoding.

```{r}
df_encoded <- df %>%
  mutate(across(ocean_proximity, as.factor)) %>%
  mutate(ocean_proximity = as.factor(ocean_proximity))

ocean_dummies <- model.matrix(~ ocean_proximity - 1, data = df_encoded)
colnames(ocean_dummies) <- paste0("ocean_", levels(df_encoded$ocean_proximity))

df_encoded <- cbind(df_encoded, as.data.frame(ocean_dummies))

cat("Columns after encoding:\n")
colnames(df_encoded)
```

---

## 3. Data Visualization

### 3.1 Histogram of Housing Median Age

Let's start with a histogram to understand the distribution of the `housing_median_age`.

```{r}
ggplot(df, aes(x = housing_median_age)) +
  geom_histogram(bins = 20, color = "black", fill = "lightblue") +
  labs(
    title = "Distribution of Housing Median Age",
    x = "Housing Median Age",
    y = "Frequency"
  ) +
  theme_minimal()
```

### 3.2 Scatter Plot: Median Income vs. Median House Value

A scatter plot can be useful to observe potential relationships between `median_income` and `median_house_value`.

```{r}
ggplot(df, aes(x = median_income, y = median_house_value)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Median Income vs. Median House Value",
    x = "Median Income",
    y = "Median House Value"
  ) +
  theme_minimal()
```

### 3.3 Geographical Plot: Locations by House Value

For a more geographic perspective, you can create a scatter plot of the properties using `longitude` and `latitude`. You might color the points by `median_house_value`.

```{r}
ggplot(df, aes(x = longitude, y = latitude, color = median_house_value)) +
  geom_point(alpha = 0.6) +
  scale_color_viridis_c() +
  labs(
    title = "Geographical Distribution of House Values",
    x = "Longitude",
    y = "Latitude",
    color = "Median House Value"
  ) +
  theme_minimal()
```

---

## 4. Additional Analysis Ideas

- **Relationship Analysis:** Consider plotting total rooms vs. population, or households vs. population.
    
- **Box Plots:** To understand the distribution within groups, use box plots for variables segmented by `ocean_proximity`.
    
- **Correlation Matrix:** Generate a heatmap to visualize the correlation between different numerical variables.
    

```{r}
numeric_df <- df %>% select(where(is.numeric))
corr_matrix <- cor(numeric_df, use = "complete.obs")
cat("Correlation matrix:\n")
print(corr_matrix)

library(corrplot)
corrplot(corr_matrix, method = "circle")
```


## 5. Machine learning analysis

### 5.1 Linear Regression Model

```{r}

# Linear regression using key numerical predictors
lm_model <- lm(median_house_value ~ median_income + housing_median_age + total_rooms + total_bedrooms + population + households, data = ml_df)

# View summary of the linear regression model
summary(lm_model)

# Add predictions
ml_df$predicted_lm <- predict(lm_model)

# Plot predicted vs actual
ggplot(ml_df, aes(x = predicted_lm, y = median_house_value)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(
    title = "Linear Regression: Predicted vs Actual Median House Value",
    x = "Predicted House Value",
    y = "Actual House Value"
  ) +
  theme_minimal()
```

### 5.2 Classification tree
```{r}
install.packages(rpart.plot)
library(rpart)
library(rpart.plot)

# Create a binary variable for classification
ml_df <- ml_df %>%
  mutate(high_value = ifelse(median_house_value > median(median_house_value), "High", "Low"),
         high_value = as.factor(high_value))

# Fit a classification tree
class_tree <- rpart(
  high_value ~ median_income + total_rooms + population +
    `ocean_NEAR BAY` + `ocean_<1H OCEAN` + `ocean_INLAND` + `ocean_NEAR OCEAN` + `ocean_ISLAND`,
  data = ml_df,
  method = "class"
)


# Plot the tree
rpart.plot(class_tree, type = 3, extra = 104, fallen.leaves = TRUE, main = "Classification Tree for High vs Low Value Homes")

```

### 5.3 Regression Tree
```{r}
# Fit a regression tree to predict house value
reg_tree <- rpart(median_house_value ~ median_income + housing_median_age + total_rooms + total_bedrooms + population + households, 
                  data = ml_df, method = "anova")

# Plot the regression tree
rpart.plot(reg_tree, type = 3, fallen.leaves = TRUE, main = "Regression Tree for Predicting Median House Value")

```


